---
title: "surprisal_stats_region_average"
output: html_document
date: "2026-08-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# GPT2

## Reading data

```{r}
library(tidyverse)

gpt2.object_path <- "data/model_outputs/gpt2_wilcox2022_embed4_object_gap_stimuli_with_surprisals.csv"

gpt2.subject_path <- "data/model_outputs/gpt2_wilcox2022_embed4_subject_gap_stimuli_with_surprisals.csv"

GRNN.object_path <- "data/model_outputs/grnn_wilcox2022_embed4_object_gap_stimuli_with_surprisals.csv"

GRNN.subject_path <- "data/model_outputs/grnn_wilcox2022_embed4_subject_gap_stimuli_with_surprisals.csv"

ngram.object_path <- "data/model_outputs/ngram_wilcox2022_embed4_object_gap_stimuli_with_surprisals.csv"

ngram.subject_path <- "data/model_outputs/ngram_wilcox2022_embed4_subject_gap_stimuli_with_surprisals.csv"

read_and_combine_dfs <- function(object_surprisals, subject_surprisals,  frequency_column, surprisal_column) {
  full_surprisals.object <- read_csv(object_surprisals)
  
  full_surprisals.subject <- read_csv(subject_surprisals)
  
  surprisals_df.object <- data.frame(
    "id" = as_factor(full_surprisals.object$sentence_group), 
    "wh" = as_factor(full_surprisals.object$wh), 
    "gap" = as_factor(full_surprisals.object$gap), 
    "condition" = full_surprisals.object$condition,
    "embed" = as_factor(full_surprisals.object$levels_of_embedding), 
    "surprisal" = full_surprisals.object[[surprisal_column]],
    "log_freq" = log2(full_surprisals.object[[frequency_column]]),
    "gap_type" = "object")
  
  surprisals_df.subject <- data.frame(
    "id" = as_factor(full_surprisals.subject$sentence_group), 
    "wh" = as_factor(full_surprisals.subject$wh), 
    "gap" = as_factor(full_surprisals.subject$gap), 
    "condition" = full_surprisals.subject$condition,
    "embed" = as_factor(full_surprisals.subject$levels_of_embedding), 
    "surprisal" = full_surprisals.subject[[surprisal_column]],
    "log_freq" = log2(full_surprisals.subject[[frequency_column]]),
    "gap_type" = "subject"  
  )
  
  
  output <- rbind(surprisals_df.subject, surprisals_df.object)
  
  output$gap_type <- as_factor(output$gap_type) 
  
  
  contrasts(output$gap) <- contr.sum(levels(output$gap))
  
  contrasts(output$wh) <- contr.sum(levels(output$wh))
  
  contrasts(output$gap_type) <- contr.sum(levels(output$gap_type))
  
  
  contrasts(output$embed) <- contr.treatment(n = levels(output$embed), base = 1)
  
  return (output)
  
}

local_frequency_column = "local_frequency" 
local_surprisal_column = "local_surprisal"

local_mean_frequency_column = "local_frequency_mean"
local_mean_surprisal_column = "local_surprisal_mean"
```

## Model fit

```{r}
first_token_surprisals_df.gpt2 <- read_and_combine_dfs(gpt2.object_path, gpt2.subject_path, local_frequency_column, local_surprisal_column)

mean_surprisals_df.gpt2 <- read_and_combine_dfs(gpt2.object_path, gpt2.subject_path, local_mean_frequency_column, local_mean_surprisal_column)

```

```{r}
library(lme4)

library(lmerTest)




surprisals.gpt_first_token.model <- lmer(surprisal ~ wh * gap * gap_type * embed + log_freq+ (1  | id), 
                             data = first_token_surprisals_df.gpt2)

surprisals.gpt_mean.model <- lmer(surprisal ~ wh * gap * gap_type * embed + log_freq + (1 | id), 
                                         data = mean_surprisals_df.gpt2)

summary(surprisals.gpt_first_token.model)
summary(surprisals.gpt_mean.model)
```

## Wh-effect

### prediction 1.A

```{r}

library(emmeans)
library(magrittr)

em <- emmeans(surprisals.gpt_first_token.model, pairwise ~ wh | gap)
em$contrasts
```

### Prediction 1.B - wh effect by distance

```{r}
emmeans(surprisals.gpt_mean.model, ~ wh * embed | gap) %>%
  contrast(interaction = c("pairwise", "trt.vs.ctrl"), by = "gap")
```

```{r}
em <- emmeans(surprisals.gpt_first_token.model, pairwise ~ wh | gap*embed)
em$contrasts


plot = emmip(em$contrasts, gap~embed, CIs= TRUE) + labs(x = "Embedding level", y = "wh effect estimate (surprisal)", color = "Gap")

ggsave("wh_first_gpt.png", plot = plot, 
       width = dev.size()[1], height = dev.size()[2], dpi = 300)

plot
```

```{r}

library(ggplot2)
library(dplyr)
library(emmeans)

wh_plot_two_models <- function(model1, model2, name1, name2,
                            ylab = "wh effect estimate (surprisal)",
                            palette = "Set1") {
  
  em1 <- emmeans(model1, pairwise ~ wh | gap*embed)
  em2 <- emmeans(model2, pairwise ~ wh | gap*embed)
  
  d1 <- emmip(em1$contrasts, gap ~ embed, CIs = TRUE, plotit = FALSE) %>% mutate(model = name1)
  d2 <- emmip(em2$contrasts, gap ~ embed, CIs = TRUE, plotit = FALSE) %>% mutate(model = name2)
  
  plot_df <- bind_rows(d1, d2)
  
  bars <- ggplot(plot_df, aes(x = embed, y = yvar, fill = interaction(gap, model))) +
    geom_col(position = position_dodge(width = 0.9)) +
    geom_errorbar(aes(ymin = LCL, ymax = UCL),
                  position = position_dodge(width = 0.9), width = 0.2) +
    theme_minimal() +
    labs(x = "Embedding level", y = ylab, fill = "Gap x Model")
  
  bars + scale_fill_brewer(palette = palette)
}

plot = wh_plot_two_models(surprisals.gpt_first_token.model, surprisals.gpt_mean.model, "first_token", "mean", ylab="wh effect estimate (surprisal) for GPT2")

ggsave("wh_first_v_mean_gpt.png", plot = plot, 
       width = dev.size()[1], height = dev.size()[2], dpi = 300)

plot
```

### Prediction 1C

```{r}
em <- emmeans(surprisals.gpt_mean.model, pairwise ~ wh | gap*gap_type)
em$contrasts

```

```{r}
em <- emmeans(surprisals.gpt_mean.model, pairwise ~ wh | gap*gap_type*embed)


plot = emmip(em$contrasts, gap*gap_type~embed, CIs= TRUE) + labs(x = "Embedding level", y = "wh effect estimate (surprisal)", color = "Gap")

ggsave("wh_subject_v_object_gpt_mean.png", plot = plot, 
       width = dev.size()[1], height = dev.size()[2], dpi = 300)

plot

```

## other questions

is the effect significant across all embedding levels and subject/object gaps?

```{r}
em <- emmeans(surprisals.gpt_mean.model, pairwise ~ wh | gap*gap_type*embed)

em$contrasts
```

```{r}
em <- emmeans(surprisals.gpt_mean.model, pairwise ~ wh | gap*embed*gap_type)

df = as.data.frame(em$contrasts)


select(df, c('gap_type', 'embed', 'p.value'))
```

## Filler-offloading effect

### Prediction 2.A

```{r}
em <- emmeans(surprisals.gpt_first_token.model,  pairwise ~ gap | wh)
em$contrasts
```

```{r}
em <- emmeans(surprisals.gpt_mean.model,  pairwise ~ gap | wh)
em$contrasts

```

### Prediction 2.B

```{r}
emmeans(surprisals.gpt_mean.model, ~ gap * embed | wh) %>%
  contrast(interaction = c("pairwise", "trt.vs.ctrl"), by = "wh")
```

```{r}
emmeans(surprisals.gpt_first_token.model, ~ gap * embed | wh) %>%
  contrast(interaction = c("pairwise", "trt.vs.ctrl"), by = "wh")
```

```{r}

em <- emmeans(surprisals.gpt_first_token.model, pairwise ~ gap | wh*embed)
em$contrasts


plot = emmip(em$contrasts, wh~embed, CIs= TRUE) + labs(x = "Embedding level", y = "filler-offloading effect estimate (surprisal)", color = "Wh")

ggsave("fo_first_token_gpt.png", plot = plot, 
       width = dev.size()[1], height = dev.size()[2], dpi = 300)

plot
```

```{r}
em <- emmeans(surprisals.gpt_mean.model, ~ wh*gap*embed)


emmip(em, wh*gap~embed, CIs= TRUE) + labs(x = "Embedding level", y = "estimate (surprisal)", color = "Wh x Gap")
```

```{r}

library(ggplot2)
library(dplyr)
library(emmeans)

fo_plot_two_models <- function(model1, model2, name1, name2,
                               ylab = "filler-offloading effect estimate (surprisal)",
                               palette = "Set1") {
  
  em1 <- emmeans(model1, pairwise ~ gap | wh*embed)
  em2 <- emmeans(model2, pairwise ~ gap | wh*embed)
  
  d1 <- emmip(em1$contrasts, wh ~ embed, CIs = TRUE, plotit = FALSE) %>% mutate(model = name1)
  d2 <- emmip(em2$contrasts, wh ~ embed, CIs = TRUE, plotit = FALSE) %>% mutate(model = name2)
  
  plot_df <- bind_rows(d1, d2)
  
  bars <- ggplot(plot_df, aes(x = embed, y = yvar, fill = interaction(wh, model))) +
    geom_col(position = position_dodge(width = 0.9)) +
    geom_errorbar(aes(ymin = LCL, ymax = UCL),
                  position = position_dodge(width = 0.9), width = 0.2) +
    theme_minimal() +
    labs(x = "Embedding level", y = ylab, fill = "Wh x Model")
  
  bars + scale_fill_brewer(palette = palette)
}

plot = fo_plot_two_models(surprisals.gpt_first_token.model, surprisals.gpt_mean.model, "first_token", "mean", ylab="filler-offloading effect estimate (surprisal) for GPT2")

ggsave("fo_first_v_mean_gpt.png", plot = plot, 
       width = dev.size()[1], height = dev.size()[2], dpi = 300)

plot
```

```{r}
em <- emmeans(surprisals.gpt_mean.model, pairwise ~ gap | wh*embed*gap_type)
em$contrasts


plot = emmip(em$contrasts, wh*gap_type~embed, CIs= TRUE) + labs(x = "Embedding level", y = "filler-offloading effect estimate (surprisal)", color = "Wh")

ggsave("fo_subject_v_object_gpt_mean.png", plot = plot, 
       width = dev.size()[1], height = dev.size()[2], dpi = 300)

plot
```

### Prediction 2.C

```{r}

em <- emmeans(surprisals.gpt_first_token.model, pairwise ~ gap | wh*gap_type*embed)
em$contrasts
```

```{r}
em <- emmeans(surprisals.gpt_mean.model, pairwise ~ gap | wh*gap_type*embed)
em$contrasts
```

```{r}
em <- emmeans(surprisals.gpt_mean.model, pairwise ~ gap | wh*gap_type*embed)


emmip(em$contrasts, wh*gap_type~embed, CIs= TRUE) + labs(x = "Embedding level", y = "wh effect estimate (surprisal)", color = "Wh")
```

```{r}

em <- emmeans(surprisals.gpt_first_token.model, pairwise ~ gap | wh*gap_type*embed)


emmip(em$contrasts, wh*gap_type~embed, CIs= TRUE) + labs(x = "Embedding level", y = "wh effect estimate (surprisal)", color = "Wh")
```

# GRNN

```{r}
first_token_surprisals_df.grnn = read_and_combine_dfs(GRNN.object_path, GRNN.subject_path, local_frequency_column, local_surprisal_column)

first_token_surprisals_df.grnn
```

```{r}
mean_surprisals_df.grnn <- read_and_combine_dfs(GRNN.object_path, GRNN.subject_path, local_mean_frequency_column, local_mean_surprisal_column)

mean_surprisals_df.grnn
```

## fitting the model

```{r}
library(lme4)

library(lmerTest)


surprisals.grnn_first_token.model <- lmer(surprisal ~ wh * gap * gap_type * embed + log_freq + (1  | id), 
                                   data = first_token_surprisals_df.grnn)


surprisals.grnn_mean.model <- lmer(surprisal ~ wh * gap * gap_type * embed + log_freq + (1 | id), 
                             data = mean_surprisals_df.grnn)


summary(surprisals.grnn_first_token.model)

summary(surprisals.grnn_mean.model)
```

### prediction 1A

```{r}
library(emmeans)
library(magrittr)

em <- emmeans(surprisals.grnn_mean.model, pairwise ~ wh | gap)
em$contrasts
```

### Prediction 1.B

```{r}
emmeans(surprisals.grnn_mean.model, ~ wh * embed | gap) %>%
  contrast(interaction = c("pairwise", "trt.vs.ctrl"), by = "gap")
```

```{r}
em <- emmeans(surprisals.grnn_mean.model, pairwise ~ wh | gap*embed)
em$contrasts


emmip(em$contrasts, gap~embed, CIs= TRUE) + labs(x = "Embedding level", y = "wh effect estimate (surprisal)", color = "Gap")
```

```{r}

plot = plot_two_models(surprisals.grnn_first_token.model, surprisals.grnn_mean.model, "first_token", "mean", ylab="wh effect estimate (surprisal) for GRNN")

ggsave("wh_first_v_mean_grnn.png", plot = plot, 
       width = dev.size()[1], height = dev.size()[2], dpi = 300)

plot
```

### prediction 1C

```{r}

em <- emmeans(surprisals.grnn_first_token.model, pairwise ~ wh | gap*gap_type)
em$contrasts
```

```{r}
em <- emmeans(surprisals.grnn_first_token.model, pairwise ~ wh | gap*gap_type*embed)



plot = emmip(em$contrasts, gap*gap_type~embed, CIs= TRUE) + labs(x = "Embedding level", y = "wh effect estimate (surprisal)", color = "Gap")

ggsave("wh_subject_v_object_grnn_first_token.png", plot = plot, 
       width = dev.size()[1], height = dev.size()[2], dpi = 300)

plot
```

### misc

```{r}
emmeans(surprisals.grnn_mean.model, ~ wh | gap * embed * gap_type) %>%
  contrast(interaction = c("pairwise", "pairwise"), by = c("gap", "embed", "gap_type"))

```

## Filler-offloading effect

## 2.A

```{r}
em <- emmeans(surprisals.grnn_first_token.model,  pairwise ~ gap | wh)
em$contrasts
```

```{r}
em <- emmeans(surprisals.grnn_mean.model,  pairwise ~ gap | wh)
em$contrasts
```

### prediction 2.B

```{r}
emmeans(surprisals.grnn_first_token.model, ~ gap * embed | wh) %>%
  contrast(interaction = c("pairwise", "trt.vs.ctrl"), by = "wh")
```

```{r}
emmeans(surprisals.grnn_mean.model, ~ gap * embed | wh) %>%
  contrast(interaction = c("pairwise", "trt.vs.ctrl"), by = "wh")
```

```{r}
em <- emmeans(surprisals.grnn_first_token.model, pairwise ~ gap | wh*embed)
em$contrasts


emmip(em$contrasts, wh~embed, CIs= TRUE) + labs(x = "Embedding level", y = "wh effect estimate (surprisal)", color = "Wh")
```

```{r}
em <- emmeans(surprisals.gpt_mean.model, pairwise ~ gap | wh*embed)
em$contrasts


emmip(em$contrasts, wh~embed, CIs= TRUE) + labs(x = "Embedding level", y = "wh effect estimate (surprisal)", color = "Wh")
```

```{r}
plot = fo_plot_two_models(surprisals.grnn_first_token.model, surprisals.grnn_mean.model, "first_token", "mean", ylab="filler-offloading effect estimate (surprisal) for GPT2")

ggsave("fo_first_v_mean_grnn.png", plot = plot, 
       width = dev.size()[1], height = dev.size()[2], dpi = 300)

plot
```

```{r}

```

### prediction 2C

```{r}
em <- emmeans(surprisals.grnn_first_token.model, pairwise ~ gap | wh*gap_type*embed)
em$contrasts
```

# n-gram data

### read data + fit model

```{r}
first_token_surprisals_df.ngram <- read_and_combine_dfs(ngram.object_path, ngram.subject_path, local_frequency_column, local_surprisal_column)


mean_surprisals_df.ngram <- read_and_combine_dfs(ngram.object_path, ngram.subject_path, local_mean_frequency_column, local_mean_surprisal_column)


first_token_surprisals_df.ngram
mean_surprisals_df.ngram
```

```{r}
library(lme4)

library(lmerTest)

surprisals.ngram_first_token.model <- lmer(surprisal ~ wh * gap * gap_type * embed + log_freq + (1 + wh*gap | id), 
                              data = first_token_surprisals_df.ngram)



surprisals.ngram_mean.model <- lmer(surprisal ~ wh * gap * gap_type * embed + log_freq + (1 + wh*gap | id), 
                                           data = mean_surprisals_df.ngram)

summary(surprisals.ngram_first_token.model)

summary(surprisals.ngram_mean.model)

```

### prediction 1.A

```{r}
library(emmeans)
library(magrittr)

em <- emmeans(surprisals.ngram.model, pairwise ~ wh | gap)
em$contrasts
```

### prediction 1.B

```{r}
emmeans(surprisals.ngram.model, ~ wh * embed | gap) %>%
  contrast(interaction = c("pairwise", "trt.vs.ctrl"), by = "gap")
```

```{r}
em <- emmeans(surprisals.ngram.model, pairwise ~ wh | gap*embed)
em$contrasts


emmip(em$contrasts, gap~embed, CIs= TRUE) + labs(x = "Embedding level", y = "wh effect estimate (surprisal)", color = "Gap")
```

```{r}
plot_two_models(surprisals.ngram_first_token.model, surprisals.ngram_mean.model, "first_token", "mean", ylab="wh effect estimate (surprisal) for ngram")
```

### prediction 1.C

```{r}

em <- emmeans(surprisals.ngram.model, pairwise ~ wh | gap*gap_type)
em$contrasts
```

```{r}

emmeans(surprisals.ngram_mean.model, ~ wh * gap_type | gap * embed) %>%
  contrast(interaction = c("pairwise", "pairwise"), by = c("gap", "embed"))


em <- emmeans(surprisals.ngram_first_token.model, pairwise ~ wh | gap*gap_type*embed)



plot = emmip(em$contrasts, gap*gap_type~embed, CIs= TRUE) + labs(x = "Embedding level", y = "wh effect estimate (surprisal)", color = "Wh")

ggsave("wh_subject_v_object_ngram_first_token.png", plot = plot, 
       width = dev.size()[1], height = dev.size()[2], dpi = 300)

plot
```

## filler-offloading effect

```{r}


em <- emmeans(surprisals.ngram_mean.model, pairwise ~ gap | wh*embed*gap_type)
em$contrasts


plot = emmip(em$contrasts, wh*gap_type~embed, CIs= TRUE) + labs(x = "Embedding level", y = "filler-offloading effect estimate (surprisal)", color = "Wh")

ggsave("fo_subject_v_object_ngram_first_token.png", plot = plot, 
       width = dev.size()[1], height = dev.size()[2], dpi = 300)

plot
```

# More visuals

```{r}
library(ggplot2)
library(dplyr)
library(emmeans)

wh_plot_three_models <- function(model1, model2, model3, name1, name2, name3,
                              ylab = "wh effect estimate (surprisal)",
                              palette = "Set1") {
  
  em1 <- emmeans(model1, pairwise ~ wh | gap*embed)
  em2 <- emmeans(model2, pairwise ~ wh | gap*embed)
  em3 <- emmeans(model3, pairwise ~ wh | gap*embed)
  
  d1 <- emmip(em1$contrasts, gap ~ embed, CIs = TRUE, plotit = FALSE) %>% mutate(model = name1)
  d2 <- emmip(em2$contrasts, gap ~ embed, CIs = TRUE, plotit = FALSE) %>% mutate(model = name2)
  d3 <- emmip(em3$contrasts, gap ~ embed, CIs = TRUE, plotit = FALSE) %>% mutate(model = name3)
  
  plot_df <- bind_rows(d1, d2, d3)
  
  bars <- ggplot(plot_df, aes(x = embed, y = yvar, fill = interaction(gap, model))) +
    geom_col(position = position_dodge(width = 0.9)) +
    geom_errorbar(aes(ymin = LCL, ymax = UCL),
                  position = position_dodge(width = 0.9), width = 0.2) +
    theme_minimal() +
    labs(x = "Embedding level", y = ylab, fill = "Gap x Model")
  
  bars + scale_fill_brewer(palette = palette)
}
```

```{r}
plot = wh_plot_three_models(surprisals.gpt_mean.model, surprisals.grnn_mean.model, surprisals.ngram_mean.model, "GPT2", "GRNN", "NGRAM", ylab="wh effect estimate (surprisal)")

ggsave("wh_effect_plot_all_models_mean_bars.png", plot = plot, 
       width = dev.size()[1], height = dev.size()[2], dpi = 300)

```

```{r}
fo_plot_three_models <- function(model1, model2, model3, name1, name2, name3,
                                 ylab = "filler-offloading effect estimate (surprisal)",
                                 palette = "Set1") {
  
  em1 <- emmeans(model1, pairwise ~ gap | wh*embed)
  em2 <- emmeans(model2, pairwise ~ gap | wh*embed)
  em3 <- emmeans(model3, pairwise ~ gap | wh*embed)
  
  d1 <- emmip(em1$contrasts, wh ~ embed, CIs = TRUE, plotit = FALSE) %>% mutate(model = name1)
  d2 <- emmip(em2$contrasts, wh ~ embed, CIs = TRUE, plotit = FALSE) %>% mutate(model = name2)
  d3 <- emmip(em3$contrasts, wh ~ embed, CIs = TRUE, plotit = FALSE) %>% mutate(model = name3)
  
  plot_df <- bind_rows(d1, d2, d3)
  
  bars <- ggplot(plot_df, aes(x = embed, y = yvar, fill = interaction(wh, model))) +
    geom_col(position = position_dodge(width = 0.9)) +
    geom_errorbar(aes(ymin = LCL, ymax = UCL),
                  position = position_dodge(width = 0.9), width = 0.2) +
    theme_minimal() +
    labs(x = "Embedding level", y = ylab, fill = "Gap x Model")
  
  bars + scale_fill_brewer(palette = palette)
}
```

```{r}
plot = fo_plot_three_models(surprisals.gpt_mean.model, surprisals.grnn_mean.model, surprisals.ngram_mean.model, "GPT2", "GRNN", "NGRAM", ylab="filler-offloading effect estimate (surprisal)")

ggsave("fo_effect_plot_all_models_mean_bars.png", plot = plot, 
       width = dev.size()[1], height = dev.size()[2], dpi = 300)

plot
```

```{r}
plot_two_models_by_type <- function(model1, model2, name1, name2,
                            ylab = "wh effect estimate (surprisal)",
                            palette = "Set1") {
  
  em1 <- emmeans(model1, pairwise ~ wh | gap*embed*gap_type)
  em2 <- emmeans(model2, pairwise ~ wh | gap*embed*gap_type)
  
  d1 <- emmip(em1$contrasts, gap ~ embed*gap_type, CIs = TRUE, plotit = FALSE) %>% mutate(model = name1)
  d2 <- emmip(em2$contrasts, gap ~ embed*gap_type, CIs = TRUE, plotit = FALSE) %>% mutate(model = name2)
  
  plot_df <- bind_rows(d1, d2)
  plot_df <- plot_df %>% filter(embed == 0)
  
  bars <- ggplot(plot_df, aes(x = embed, y = yvar, fill = interaction(gap, model, gap_type))) +
    geom_col(position = position_dodge(width = 0.9)) +
    geom_errorbar(aes(ymin = LCL, ymax = UCL),
                  position = position_dodge(width = 0.9), width = 0.2) +
    theme_minimal() +
    labs(x = "Embedding level", y = ylab, fill = "Gap x Model x Gap_type")
  
  bars + scale_fill_brewer(palette = palette)
}
```

```{r}
plot_two_models_by_type(surprisals.ngram_first_token.model, surprisals.ngram_mean.model, "first_token", "mean", ylab="wh effect estimate (surprisal) for ngram")
```
