---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

## Computational modelling of filler-gap dependencies

```{r}

```

```{r}

```

## GPT2: mixed effect model of the effects of presence of wh-licensor, presence of gap, gap type and embedding level on surprisals

```{r}
library(tidyverse)

gpt2.object_path <- "data/model_outputs/gpt2_wilcox2022_embed4_object_gap_stimuli_with_surprisals.csv"

gpt2.subject_path <- "data/model_outputs/gpt2_wilcox2022_embed4_subject_gap_stimuli_with_surprisals.csv"

read_and_combine_dfs <- function(object_surprisals, subject_surprisals) {
  full_surprisals.object <- read_csv(object_surprisals)
  
  full_surprisals.subject <- read_csv(subject_surprisals)
  
  surprisals_df.object <- data.frame(
    "id" = as_factor(full_surprisals.object$sentence_group), 
    "wh" = as_factor(full_surprisals.object$wh), 
    "gap" = as_factor(full_surprisals.object$gap), 
    "condition" = full_surprisals.object$condition,
    "embed" = as_factor(full_surprisals.object$levels_of_embedding), 
    "surprisal" = full_surprisals.object$local_surprisal,
    "gap_type" = "object")
  
  surprisals_df.subject <- data.frame(
    "id" = as_factor(full_surprisals.subject$sentence_group), 
    "wh" = as_factor(full_surprisals.subject$wh), 
    "gap" = as_factor(full_surprisals.subject$gap), 
    "condition" = full_surprisals.subject$condition,
    "embed" = as_factor(full_surprisals.subject$levels_of_embedding), 
    "surprisal" = full_surprisals.subject$local_surprisal,
    "gap_type" = "subject"  
  )
  
  
  output <- rbind(surprisals_df.subject, surprisals_df.object)
  
  output$gap_type <- as_factor(output$gap_type) 
  
  
  contrasts(output$gap) <- contr.sum(levels(output$gap))
  
  contrasts(output$wh) <- contr.sum(levels(output$wh))

  contrasts(output$gap_type) <- contr.sum(levels(output$gap_type))

  
  contrasts(output$embed) <- contr.treatment(n = levels(output$embed), base = 1)

  return (output)
  
}



surprisals_df.gpt2 <- read_and_combine_dfs(gpt2.object_path, gpt2.subject_path)

head(surprisals_df.gpt2)
summary(surprisals_df.gpt2)

contrasts(surprisals_df.gpt2$gap)
contrasts(surprisals_df.gpt2$wh)
contrasts(surprisals_df.gpt2$gap_type)
contrasts(surprisals_df.gpt2$embed)

```

## Fitting a Mixed-effects model to the gpt2 data

**TODO**: make all of this be functions so I can reuse between models!

```{r}

library(lme4)

library(lmerTest)

surprisals.gpt.model <- lmer(surprisal ~ wh * gap * gap_type * embed + (1 + wh*gap | id), 
           data = surprisals_df.gpt2)

summary(surprisals.gpt.model)
```

```{r}

plot(fitted(surprisals.gpt.model), residuals(surprisals.gpt.model))

```

## Analysis with emmeans package

### break down results by  gap type 

```{r}
library(emmeans)
library(magrittr)

em.gap_type <- emmeans(surprisals.gpt.model, specs =  ~ wh*gap | embed)
em.gap <- emmeans(surprisals.gpt.model, spec = ~ wh | gap)
em.wh <- emmeans(surprisals.gpt.model, spec = ~ gap*embed| wh)
em.gap_type <- emmeans(surprisals.gpt.model, spec = ~ gap | gap_type)
```

on average sentences with subject-gaps have higher s

```{r}
joint_tests(em.wh, by="wh")
```

urprisal

```{r}
emmeans(surprisals.gpt.model, spec = ~ gap_type) %>% plot

```

```{r}
summary(em.gap_type)
```

wh alone only a significant predictor for object gaps – I think this makes sense with the theory that LMs are bad at subject-gaps

embedding alone only a predictor for subject gaps, and wh:embed interaction is also only present for subject gaps

```{r}

joint_tests(surprisals.gpt.model, by="embed")
```

This plot seems broadly pretty useful? But I don't know if it tells us that much because it's just means – and we generally care a lot more about differences in surprisal between conditions for the same sentence?

```{r}
emmip(surprisals.gpt.model, gap_type ~ embed | wh*gap)
```

## Further break down by embedding level

Again significant for subject gaps — across all embedding levels – but not object gaps

```{r}
emmeans(surprisals.gpt.model, trt.vs.ctrl ~ embed | gap_type)
```

Breaking down even more – significant for wh=yes and gap_type = subject, not anything else

```{r}
emmeans(surprisals.gpt.model, trt.vs.ctrl ~ embed | gap_type * wh)
```

Embedding consistently has an effect on outcomes for +wh_gap condition across both gap types. Bigger effect for higher embedding levels for object but not subject gaps – makes sense if we think LMs are just generally bad at subject gaps maybe. For gap=no, wh= yes effect present at **high embedding levels for object** and **low embedding levels for subject**

```{r}
emmeans(surprisals.gpt.model, trt.vs.ctrl ~ embed | gap_type * gap * wh)

```

## Examining the wh-effect

```{r}
emmeans(surprisals.gpt.model, pairwise ~ wh | gap)
```

```{r}
emmeans(surprisals.gpt.model, pairwise ~ wh | gap * gap_type)
```

## Examining the filler-offloading effect: statistically significant difference but NOT in the direction we expect

something makes gaps always more surprising regardless of their grammaticality! This is stronger for subject-gaps meaning that they are less good at filler-offloading for subject gaps

```{r}
emmeans(surprisals.gpt.model, pairwise ~ gap | wh)
```

```{r}
emmeans(surprisals.gpt.model, pairwise ~ gap | wh * gap_type)
```

## GRNN: mixed effect model of the effects of presence of wh-licensor, presence of gap, gap type and embedding level on surprisals from

```{r}
library(tidyverse)

grnn.object_path = "data/model_outputs/grnn_wilcox2022_embed4_object_gap_stimuli_with_surprisals.csv"

grnn.subject_path = "data/model_outputs/grnn_wilcox2022_embed4_subject_gap_stimuli_with_surprisals.csv"


surprisals_df.grnn <- read_and_combine_dfs(grnn.object_path, grnn.subject_path)

head(surprisals_df.grnn)
summary(surprisals_df.grnn)


```

## Fitting a mixed-effects linear model to GRNN surprisal data

```{r}
surprisals.grnn.model <- lmer(surprisal ~ wh * gap * gap_type * embed + (1 + wh*gap | id), 
                             data = surprisals_df.grnn)

summary(surprisals.grnn.model)
```

```{r}
plot(fitted(surprisals.gpt.model), residuals(surprisals.gpt.model))

```

## Analysis with emmeans package

### break down results by  gap type 

```{r}
library(emmeans) 
library(magrittr)  
em.grnn.gap_type <- emmeans(surprisals.grnn.model, specs =  ~ wh*gap*embed | gap_type)
```

on average sentences with subject-gaps have higher surprisal

```{r}
emmeans(surprisals.grnn.model, spec = ~ gap_type) %>% plot 
```

```{r}
summary(em.grnn.gap_type)
```

wh alone not a significant predictor – unlike gpt2

embedding alone only a predictor for subject gaps, and wh:embed interaction is also only present for subject gaps: again maybe this aligns with the theory that the models are just really really bad when it comes to subject gaps?

```{r}
joint_tests(surprisals.grnn.model, by="gap_type")
```

```{r}
emmip(surprisals.gpt.model, gap_type ~ embed | wh*gap)
```

## 

## Further break down by embedding level

Again significant for subject gaps — across all embedding levels – but not object gaps

```{r}
emmeans(surprisals.grnn.model, trt.vs.ctrl ~ embed | gap_type)
```

Breaking down even more – significant for wh=yes and gap_type = subject, not anything else

```{r}
emmeans(surprisals.grnn.model, trt.vs.ctrl ~ embed | gap_type * wh)
```

Significant effects only for wh_gap condition across both gap types, no significant interaction in the ungrammatical wh_no_gap condition – GRNN more superficial in its modelling?

```{r}
emmeans(surprisals.grnn.model, trt.vs.ctrl ~ embed | gap_type * gap * wh) 
```

## 

## Examining the wh-effect

```{r}
emmeans(surprisals.grnn.model, pairwise ~ wh | gap)
```

```{r}
emmeans(surprisals.grnn.model, pairwise ~ wh | gap * gap_type)
```

## Examining the filler-offloading effect: statistically significant difference but NOT in the direction we expect

something makes gaps always more surprising regardless of their grammaticality! This is stronger for subject-gaps meaning that they are less good at filler-offloading for subject gaps

```{r}
emmeans(surprisals.gpt.model, pairwise ~ gap | wh)
```

```{r}
emmeans(surprisals.gpt.model, pairwise ~ gap | wh * gap_type)
```

## 
